### 2013-03-24 doc/sosp13/abstract.tex
minor edits all over

Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. Grappa's goal is to provide scalable performance for irregular
parallel applications, such as graph processing. These applications are
challenging because they exhibit little data locality, imbalanced parallel
work and complex communication patterns.

Grappa serves both as a C++ user library and as a foundation upon which higher
level languages can be developed or adapted. Grappa tolerates delays to remote
memory by multiplexing thousands of lightweight tasks to each processor core;
balances load via fine-grained distributed work-stealing; takes full advantage
of network characteristics by aggregating smaller data requests into large
ones; and provides efficient synchronization and remote operations. We present
a detailed description of the Grappa system, programming examples using the
library interface, and compare performance for several irregular benchmarks to
hand-optimized MPI code and the Cray XMT, a custom system used to target the
real time graph analytics market.

>>>>

\Grappa is a runtime system for commodity clusters of multicore computers that
presents a massively parallel, single address space abstraction to
applications. \Grappa's goal is to provide scalable performance for irregular
parallel applications, such as graph processing. These applications are
challenging because they exhibit little data locality, imbalanced parallel
work and complex communication patterns.

\Grappa serves both as a C++ user library and as a foundation upon which
higher level languages can be developed or adapted. \Grappa tolerates delays
to remote memory by multiplexing thousands of lightweight execution contexts
to each processor core; balances load via fine-grained distributed
work-stealing; takes full advantage of network characteristics by aggregating
smaller data requests into large ones; and provides efficient synchronization
and remote operations. We present a detailed description of the \Grappa
system, programming examples using the library interface, and compare
performance for several irregular benchmarks to hand-optimized MPI code and
the Cray XMT, a custom system used to target the real time graph analytics
market.

### 2013-03-24 doc/sosp13/overview.tex
little tweaks
minor edits all over

\section{Grappa Overview}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

Grappa (Figure~\ref{fig:grappa}) has three main software components:
\begin{description}

\item [Tasking system.] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker
threads than required for latency tolerance. By having at least
several threads ready to run at all times, the scheduler can prefetch
a few stack frames into cache to lower the likelihood of costly main
memory accesses on every context switch.

@@

% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network. Therefore, Grappa can afford to \emph{trade latency for throughput}: by {\em increasing} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).

@@

In the next three sections we describe both Grappa's main capabilities and how they are implemented.

>>>>

\section{\Grappa Overview}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} \Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\Grappa (Figure~\ref{fig:grappa}) has three main software components:
\begin{description}

\item [Tasking system.] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker threads than
required for latency tolerance. By having at least several threads ready to
run at all times, the scheduler can prefetch a context data into cache to
lower the likelihood of costly main memory accesses during a context switch.

@@

% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network. Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).

@@

In the next three sections we describe both \Grappa's main components and how they are implemented.

### 2013-03-24 doc/sosp13/tasks.tex
minor edits all over

During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{ yield() } ] \hfill \\
      Yields core to scheduler, enqueuing caller to be scheduled again soon
    \item[ \texttt{ suspend() } ] \hfill \\
      Yields core to scheduler, enqueuing caller only once another task calls wake
    \item[ \texttt{ wake( task * $t$ ) } ] \hfill \\
      Enqueues some other task $t$ to be scheduled again soon
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} \Grappa scheduling API } 
    \end{minipage}
  \end{center}
\end{figure}

@@

\paragraph{Context switching} Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes --- importantly, a context is saved in a single 64 cache line.

@@

\paragraph{Work stealing} 
When the scheduler finds no work to assign to its workers, it commences to
steal tasks from other cores using an asynchronous \texttt{call\_on} active
message. It chooses a victim at random until it finds one with a non-zero
amount of work in its public task queue. The scheduler steals half of the
tasks it finds at the victim. Work stealing is particularly interesting in
Grappa since performance depends on having many active worker threads on each
core. Even if there are many active threads, if they are all suspended on
long-latency operations, then the core is underutilized. We choose to initiate
one outstanding steal whenever the ready queue is empty there are idle worker
threads.

@@

\TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized. So,
there is a choice: should the core use some of this idle time to perform
work stealing or should it just wait as it is possible that local tasks
will create more work. We choose to have a core initiate a steal when
all of the following conditions hold: no workers are ready to run, the
unstarted task queues are empty, and there are no outstanding steal
requests. Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting
tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
Termination detection is not built into the Grappa task scheduler,
rather, it is considered a programming error for the program to exit
without syncing all tasks.
}

@@

Grappa programmers focus on expressing as much
parallelism as possible without concern for where it will execute.
Grappa then chooses where and when to exploit this parallelism,
scheduling as much work as is necessary on each core to keep it busy
in the presence of system latencies and task dependences.

Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}:

First, when the programmer identifies work
that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a
\texttt{spawn}.

Second, a programmer can use \texttt{spawn\_on} to spawn a task on a
specific core in the system or at the home core of a particular memory
location.

Third, the programmer can invoke a parallel for loop, provided that the trip
count is known at loop entry. The programmer specifies a function pointer
along with start and end indices and an optional threshold to control parallel
overhead. Grappa does {\em recursive decomposition} of iterations, similar to
Cilk's cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold.

Fourth, a programmer may want to execute an active message; that is,
to run a small piece of code on a particular core in the system
without waiting for execution resources to be available.  For example,
custom synchronization primitives execute this way, as a function
executed on the core where the data lives.  Grappa provides the
\texttt{call\_on} call for this purpose.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} Grappa API: expressing parallelism
      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
      } % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

>>>>

During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the \Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

\begin{figure}[htbp]
  \begin{center}
	\begin{tabular}{l}
    \texttt{\scriptsize yield() } \\
      Yields core to scheduler, enqueuing caller to be \\ scheduled again soon \\
    \texttt{\scriptsize suspend() }  \\
      Yields core to scheduler, enqueuing caller only once \\ another task calls wake \\
    \texttt{\scriptsize wake( task * $t$ ) } \\
      Enqueues some other task $t$ to be scheduled again soon \\
	\end{tabular}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:scheduling} \Grappa scheduling API } 
    \end{minipage}
  \end{center}
\end{figure}

@@

\paragraph{Context switching} \Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes --- importantly, a context is saved in a single 64 cache line.

@@

\paragraph{Work stealing} 
When the scheduler finds no work to assign to its workers, it commences to
steal tasks from other cores using an asynchronous \texttt{call\_on} active
message. It chooses a victim at random until it finds one with a non-zero
amount of work in its public task queue. The scheduler steals half of the
tasks it finds at the victim. Work stealing is particularly interesting in
\Grappa since performance depends on having many active worker threads on each
core. Even if there are many active threads, if they are all suspended on
long-latency operations, then the core is underutilized. We choose to initiate
one outstanding steal whenever the ready queue is empty there are idle worker
threads.

@@

\TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized. So,
there is a choice: should the core use some of this idle time to perform
work stealing or should it just wait as it is possible that local tasks
will create more work. We choose to have a core initiate a steal when
all of the following conditions hold: no workers are ready to run, the
unstarted task queues are empty, and there are no outstanding steal
requests. Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting
tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
Termination detection is not built into the \Grappa task scheduler,
rather, it is considered a programming error for the program to exit
without syncing all tasks.
}

@@

\Grappa programmers focus on expressing as much parallelism as possible
without concern for where it will execute. \Grappa then chooses where and when
to exploit this parallelism, scheduling as much work as is necessary on each
core to keep it busy in the presence of system latencies and task dependences.

\Grappa provides four methods for expressing parallelism, shown in
Figure~\ref{fig:expressing-parallelism}. First, when the programmer identifies
work that can be done in parallel, the work may be wrapped up in a function
and queued with its arguments for later execution using a \texttt{spawn}.
Second, a programmer can use \texttt{spawn\_on} to spawn a task on a specific
core in the system or at the home core of a particular memory location. Third,
the programmer can invoke a parallel for loop, provided that the trip count is
known at loop entry. The programmer specifies a function pointer along with
start and end indices and an optional threshold to control parallel overhead.
\Grappa does {\em recursive decomposition} of iterations, similar to Cilk's
cilk\_for construct~\cite {cilkforimplementation}, and TBB's {\tt
parallel\_for}~\cite{intel_tbb}. It generates a logarithmically-deep tree of
tasks, stopping to execute the loop body when the number of iterations is
below the required threshold. Fourth, a programmer may want to execute an active message; that is, to run a
small piece of code on a particular core in the system without waiting for
execution resources to be available. For example, custom synchronization
primitives execute this way, as a function executed on the core where the data
lives. \Grappa provides the \texttt{call\_on} call for this purpose.

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      \Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
      } % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}
